Language Skills in an Ethnically Segmented Labour market: Estonia 1989 -- 2012

### 2018-07-08 paper/ijm.tex
paper: conclusion rewritten

We document that different languages are associated in a
different way with unemployment and wage. Our most
intriguing observation is that
Russian men earn virtually no income premium from being fluent in Estonian,
the majority and official language of the country, while the fluency is related to a substantially
lower unemployment rate.
Our analysis supports \citet{YaoandOurs2015} and \citet{Lindemann2013} finding that
the labour market is segmented between ethnic and gender lines.
However, we find that ethnic segregation is less prevalent at the lower end
of the male labour market.
The lower--end labour market, associated with frequent unemployment, is
better integrated and
those who are fluent in the language may expect both less unemployment
and higher wage.  However, men at better paid and more stable jobs
gain little from Estonian.  We
suppose this is because the middle and upper-end labour market is more
segregated and even
those who speak the official language cannot improve their income
much.  This is possibly a type of glass-ceiling effect.  This hypothesis is
supported by \citet{leppik+vihalemm2015JofBaltStud} who show that
Estonian language skills are not associated with improved career
mobility over time.  It is also corroborated with the quantile regression evidence that
Estonian language is related to a positive wage premium only at the
lowest end of income scale \citep{Toomet2011}.  There is also
anecdotal evidence suggesting that offices are
typically manned by ethnic Estonians while production units have a
large share of Russian workers.
One can speculate that this outcome is related to a substantial population share
with negative attitudes toward Russians \citep{korts2009JofBaltStud}.

The ethnic segregation seems to be less of an issue for women.
Unfortunately, neither \citet{Toomet2011} nor
\citet{leppik+vihalemm2015JofBaltStud} analyse men and women
separately.  We suppose that women are more frequently employed in jobs in which they need Estonian
language more, such as working in a direct contact with customers.
We also find evidence that over time Estonian is becoming more
important in the female labour market.  Unfortunately, the current data
does not let us to assess whether is due to decreasing segregation, or
economy--wide rise of Estonian language.  No such trend is visible for men.

The other results are more in line with other studies.
English language is related to a substantial income premium but
virtually to no effect on unemployment.  This is similar to findings by
\citet{ginsburgh+prieto-rodriguez2011ILRR} and
\citet{beblavy+2016CEPS} who show that English is more important at better paid
jobs.  In contrast, our outcomes suggest that English has little use
at low-paid, unstable jobs \citep{Toomet2011}.

Note also that for
English, we are able to explain a substantial part of the raw effect
by individual characteristics, suggesting that the true causal effect
may be smaller than our estimates.

The study has a number of limitations.  Although we believe that our
estimates are good indicators of a causal relationship, we cannot make
such claim unless a suitable instruments for language skills can be
found.  In particular, as individual characteristics are able to explain a
large part of English returns, we suppose part of the remaining effect
may also related to an ability bias.  Another weak point in the current
data, although it is
standard in the literature, is using the self-reported language
skills.  As language test data becomes available, we may be able to
decrease the measurement error in the future.

>>>>

We document that all analyzed language skills are associated with
unemployment and wage, however in a different way. Our most
intriguing observation is that the largest group of
Russian men, high-school graduates, earn virtually no income premium if fluent in Estonian,
the majority and the official language of the country, while ability
to speak the language is related to a substantially
lower unemployment rate.
Our analysis supports \citet{YaoandOurs2015}, and \citet{Lindemann2013} findings that
the labour market is segmented between ethnic and gender lines.
However, we find that ethnic segregation is less prevalent in the male
labour market.
We suspect that for majority of Russian speaking men, Estonian
language gives a distinct advantage by facilitating hiring by
Estonian-dominated employers.  However, these
jobs are associated with little career mobility.
(\citet{leppik+vihalemm2015JofBaltStud} reach a similar conclusion
based on Immigrant Population Survey).  There is also
anecdotal evidence suggesting that offices are
typically manned by ethnic Estonians while manufacturing units have a
large share of Russian workers.
One can speculate that this outcome is related to a substantial population share
with negative attitudes toward Russians \citep{korts2009JofBaltStud}.
On a more positive tone, we find
that this may not be the case for highly educated workers.

The ethnic segregation seems to be less of an issue for women.
Unfortunately, neither \citet{Toomet2011} nor
\citet{leppik+vihalemm2015JofBaltStud} analyse men and women
separately.  We find some evidence that female labor market is less
segregated along ethnic lines, and women are more frequently employed in jobs in which they need Estonian
language more, such as working in a direct contact with customers.

We also find evidence that over time Estonian is becoming more
important in the female labour market.  Unfortunately, the current data
does not let us to assess whether is due to decreasing segregation, or
economy--wide rise of the importance of Estonian language.  No such trend is visible for
men suggesting that 20 years--almost a generation after the collapse of Soviet Union, the
labor market integration is still not complete.

The other results are more in line with other studies.
English language is related to a substantial income premium but
virtually to no effect on unemployment.  This is similar to findings by
\citet{ginsburgh+prieto-rodriguez2011ILRR} and
\citet{beblavy+2016CEPS} who show that English is more important at better paid
jobs.  In contrast, our outcomes suggest that English has little use
at unstable jobs, at these jobs where workers frequently move between
employment and unemployment.
Note also that for
English, we are able to explain a substantial part of the raw effect
by individual characteristics, suggesting that the true causal effect
may be smaller than our estimates.

The study has a number of limitations.  Although we believe that our
estimates are good indicators of a causal relationship, we cannot 
claim this as our data do not contain strong enough instruments.
In particular, as individual characteristics are able to explain a
large part of English returns, we suppose part of the remaining effect
may also related to an ability bias.  Another weak point in the current
data is using the self-reported language
skills, although this is the standard practice in the literature.  As language test data becomes available, we may be able to
decrease the measurement error in the future.
